\dnest~Manual

### 2016-01-14 doc/manual.tex
Finished manual

\begin{figure}[h!]
\begin{framed}
\begin{verbatim}
$ ./main
# Using 1 thread.
# Target compression factor between levels = 2.7182818284590451
# Seeding random number generator with 1402801184.
# Thread 1: Generated 1 particles from the prior.
# Creating level 1 with logL = -51.65058877.
# Creating level 2 with logL = -37.67550971.
# Creating level 3 with logL = -28.69668897.
# Creating level 4 with logL = -21.61338509.
# Creating level 5 with logL = -15.92331209.
# Creating level 6 with logL = -11.291182.
# Saving a particle to disk. N = 1.
# Creating level 7 with logL = -7.301954778.
# Creating level 8 with logL = -4.116771404.
# Creating level 9 with logL = -1.19393518.
# Creating level 10 with logL = 1.620392784.
# Saving a particle to disk. N = 2.
\end{verbatim}
\end{framed}
\caption{\it An example of \dnest~output that is shown on the screen. More
detailed information is saved to the three output files
{\tt sample.txt}, {\tt sample\_info.txt}, and {\tt levels.txt}.
\label{fig:output}}
\end{figure}

@@

The plots are useful for diagnosing the progress of your run. For example, you
want to ensure that the Metropolis acceptance fraction doesn't drop to near
zero. If it does, the proposal distributions might be too aggressive, or maybe
you just created too many levels (and the high level distributions are just
extremely narrow). The other key plot for diagnostics is the posterior
weights. On most problems, these will peak at some point, and begin to
come back down towards zero (See Figure~\ref{fig3}).
Once this occurs, you probably have enough levels.
It is wasteful to create large amounts of levels extending well beyond the peak.
However, on problems with phase transitions more than one peak may occur.
An example of not having enough levels is given in
Figure~\ref{fig:not_enough_levels}. To change the number of levels, you
can use the OPTIONS file, described in Section~\ref{sec:options}.\\

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{fig1.pdf}
\caption{\it Level vs. Time from a DNest3 run with 100 levels.\label{fig:fig1}}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{framed}
\begin{verbatim}
DNest3 Command Line Options:
-h: display this message
-l <filename>: load level structure from the specified file.
-o <filename>: load DNest3 options from the specified file. Default=OPTIONS
-s <seed>: seed the random number generator with the specified value.
           If unspecified, the system time is used.
-d <filename>: Load data from the specified file, if required.
-c <value>: Specify a compression value (between levels) other than e.
-t <num_threads>: run on the specified number of threads. Default=1.
-f <filename>: a custom configuration file for adding problem specific
           options if required.
-z: the output "sample.txt" file will be gzipped and have the suffix ".gz".
\end{verbatim}
\end{framed}
\end{figure}

The most important option is the one about the number of threads. Depending on
your computer, you probably want to run \dnest~with more than one thread.
Each thread will be responsible for evolving its own walker(s), and periodically
the threads will pool their results (for example, information about the
levels). To execute DNest with 8 threads, simply use {\tt ./main -t 8}. You can
replace 8 with however many threads you want. I usually use 8 threads because
that's how many my computer can run simultaneously at full speed.\\

@@

The {\tt -z} option is only available is you have compiled with the {\tt zlib}
library and Boost {\tt iostreams} library installed. \\

If you have any additional configuration information required by the run, e.g.\
for setting values needed for a particular model that you don't want to hardcode,
then they can be added through specifying a configuration file using the {\tt -f}
option. You will have to write your own function to read in this file, so it can
be in whatever format you are comfortable with (e.g.\ an {\tt ini} file, or {\tt
json} file).\\

@@

\section{Implementing models}
\dnest~currently includes four example models: Atoms, SpikeSlab, FitSine, and
HarlemShake. It also contains an empty Template example which you can use
to implement your own models.\\

To use \dnest~on a particular problem, you need to implement a C++ class
which describes the features of the model (e.g. what the parameters are,
what the prior is, and what the likelihood is). Specifically, an object
(instance) of your class represents a point in the parameter space of your
problem. The class must define several methods (functions) which can be
applied to any object. The first method is {\tt fromPrior()}, which when called,
initialises the object with parameter values from the prior distribution.
The second important method is {\tt perturb()}, which is used to propose
a change to the parameters, as in the basic Metropolis algorithm. The third
important method is {\tt logLikelihood()}, which returns (as you might expect)
the log of the likelihood function evaluated at the object's position. Next,
we have the {\tt print(std::ostream\&)} method which defines how the model
parameters are to be written to the file (i.e. which parameters are to be
printed, and in what order). Finally, there is a method {\tt description()}
which allows you to put a header at the top of the output files with information
about what is in each column. To demonstrate how these methods work, we'll go
through the code for {\tt SpikeSlab}.\\

In the SpikeSlab directory,
the relevant class is the {\tt SpikeSlab} class which is defined in the files
{\tt SpikeSlab.h} and {\tt SpikeSlab.cpp}. If you open {\tt SpikeSlab.h},
you'll see that the class has only one variable, which is a C++ vector called
{\tt params}. This contains the 20 parameter values.\\

\begin{figure}
\begin{framed}
\begin{verbatim}
class SpikeSlab:public DNest3::Model
{
    private:
        std::vector<double> params;
\end{verbatim}
\end{framed}
\end{figure}

It's up to you whether you want to have a single vector of parameters, like
in {\tt SpikeSlab}, or if you want to name all your parameters with names that
make sense to you.\\

In {\tt SpikeSlab.cpp}, all of the methods of the {\tt SpikeSlab} class
are defined. First, we have the {\it constructor}, which defines what happens
when we create an object of class {\tt SpikeSlab}:\\

@@

This constructor does nothing except specify the size of the {\tt params}
vector, which in this example is 20. Next we have {\tt fromPrior()}, the
method which generates the 20 parameters from the prior. In the SpikeSlab
example, the prior is Uniform$(-0.5, 0.5)$, so this method just has a loop
which generates each value from this distribution. Here I'm making use of
a function {\tt randomU()} which generates values from $U(0, 1)$. This function
is defined in {\tt Utils.h} within \dnest, and is really just a function
wrapping a call to GSL's random number generator.\\

\begin{framed}
\begin{verbatim}
void SpikeSlab::fromPrior()
{
    for(size_t i=0; i<params.size(); i++)
        params[i] = -0.5 + randomU();
}
\end{verbatim}
\end{framed}

The next important method is {\tt perturb()}, which defines the Metropolis
proposals that are used to explore. Since \dnest~spends some time sampling
the prior, and some time sampling narrow distributions like the posterior,
it's a good idea to use very heavy tailed proposal distributions. In the
SpikeSlab example, the prior width is 1, so you shouldn't make moves much
bigger than that, but you don't have any real idea how narrow the posterior
distribution is until you've solved the problem. In the SpikeSlab example,
the proposal is done by choosing one of the 20 parameters to move, and making
a move according to a scale mixture of gaussians (equivalent to ``a gaussian
proposal with a random step size''). Finally, to prevent any proposals outside
the prior range, I use modulo to ensure the proposed value is between -0.5
and 0.5. The resulting proposal is:\\

\begin{framed}
\begin{verbatim}
double SpikeSlab::perturb()
{
    int which = randInt(params.size());
    params[which] += pow(10., 1.5-6*randomU())*randn();
    params[which] = mod(params[which] + 0.5, 1.) - 0.5;
    return 0.;
}
\end{verbatim}
\end{framed}

For convenience, I recently defined two useful functions {\tt randh()}
(to generate heavy tailed proposals) and
{\tt wrap(double\& value, double min, double max)} (to simplify the modulo
stuff). An equivalent version of the {\tt perturb} method is then:\\

\begin{framed}
\begin{verbatim}
double SpikeSlab::perturb()
{
    int which = randInt(params.size());
    params[which] += randh();
    wrap(params[which], -0.5, 0.5);
    return 0.;
}
\end{verbatim}
\end{framed}

The last thing we need to pay attention to in the perturb method is the
returned value. \dnest~assumes that your perturb method, if used repeatedly,
would inherently explore the prior distribution in your problem. With
a uniform prior, it's easy to ensure this. However, for a non-uniform prior,
it's not so easy. One way to do it is to work in a coordinate system where
the prior {\it is} uniform, and then transforming the parameter before using
it in the likelihood function. Another way is to return a nonzero value from
the {\tt perturb} method. Any value returned is added to the log of the
acceptance probability. So if you want to explore a non-uniform prior, you
can return the log of the ratio of the prior density at the proposed point
to the value at the original point. For concreteness, suppose you had a
parameter $a$ with a Uniform$(0,10)$ prior, and a parameter $b$ with a
Normal$(20, \sigma=5)$ prior. Then the {\tt perturb} method could be implemented
like this:\newpage

\begin{framed}
\begin{verbatim}
double MyModel::perturb()
{
    int which = randInt(2);
    double logH = 0.;

    if(which == 0)
    {
        a += 10.*randh();
        wrap(a, 0., 10.);
    }
    else
    {
        logH -= -0.5*pow((b - 20.)/5., 2);
        b += 5.*randh();
        logH += -0.5*pow((b - 20.)/5., 2);
    }

@@

The {\tt logLikelihood} method does what you'd expect: evaluates and returns
the log likelihood based on the current setting of the parameters. In the
SpikeSlab example, the likelihood is an invented mixture of two gaussians.
However, in general, your likelihood will depend on a data set. When I'm using
real data, I usually write a {\tt Data} class, an object of which represents
a data set. Since you need to access the dataset from within another class,
it's easiest to make the C++ equivalent of a ``global variable'': a single
static instance of the dataset which is accessible from everywhere. See
the FitSine example if you want to see a Data class. Note that the FitSine
example has been superseded by my other \dnest~related project, RJObject.\\

Finally, the {\tt print} method defines how your parameters are printed to the
{\tt sample.txt} (and subsequently {\tt posterior\_sample.txt}) files. In
SpikeSlab, the {\tt print} method writes all 20 parameters, separated by
spaces, to a single line of the stream ``out'' which is input to the method:\\

@@

Your print method needs to output parameters on a single line, in any order
you want. These will be the order in which they'll appear in the output files
{\tt sample.txt} and {\tt posterior\_sample.txt}. As an example, for our
above hypothetical model with parameters $a$ and $b$, the print function
and the description function might look like this:

@@

The Makefile in the Template directory should allow you to compile your model
by issuing the command {\tt make} in a terminal. If everything worked well
then the executable {\tt main} should exist, which you can use to run
\dnest~on your model. As always, the first time you run \dnest~on a new model,
you should pay attention to the output plots and make sure everything is okay,
such as the number of levels etc.

\subsection{Implementing a model using Julia}
Experimental support for models implemented in Julia is provided in the
{\tt Examples/JuliaModel} directory. To use this, follow these steps:

\begin{enumerate}
\item Implement your log likelihood in {\tt julia\_model.jl}, as a function
	called {\tt log\_likelihood} that takes a {\tt Vector::Float64} as an argument
	and returns a {\tt Float64}. You should assume that the priors for the
	parameters are independent Uniform(0, 1) distributions. To use
	a different prior, transform the parameters inside
	your log likelihood function before moving them.\\
\item You may add other functions to {\tt julia\_model.jl}, for example to load
	a dataset to refer to in your log likelihood.\\
\item Edit the constructor for {\tt MyModel} in {\tt MyModel.cpp} so the parameter
	vector has the correct length for your problem. The default given value
	is 20.\\
\item Open the {\tt Makefile} and make sure the {\tt -L} option passed to
	{\tt g++} points to the
	directory on your computer where the Julia shared library (such
	as {\tt libjulia.so}) is stored. You might also have to configure your
	system so that it searches for libraries here at runtime, e.g.
	by running ldconfig as root. The \dnest~library and include files should
	also be somewhere where the compiler will find them (or you can add them
	to the {\tt g++} calls in the {\tt Makefile} using {\tt -I} for header
	files and {\tt -L} for libraries).\\
\item Open {\tt main.cpp} and edit the first argument to the function
	{\tt jl\_init\_with\_image} to be the path to the file {\tt sys.so} associated with
	Julia on your system.\\
\item Compile by invoking '{\tt make}'.\\
\item Run the executable! Note: only a single thread can be used, so you might want
	to use the {\tt OPTIONS} file to have more than one particle. Five is a
	reasonable default.
\end{enumerate}

>>>>

\begin{figure}[h!]
\begin{framed}
\begin{verbatim}
$ ./main
# Using 1 thread.
# Target compression factor between levels = 2.7182818284590451
# Seeding random number generators. First seed = 1452723978.
# Generating 1 particle from the prior...done.
# Creating level 1 with log likelihood = -44.1995.
# Creating level 2 with log likelihood = -32.1028.
# Creating level 3 with log likelihood = -25.5476.
# Creating level 4 with log likelihood = -19.911.
# Creating level 5 with log likelihood = -15.5224.
# Creating level 6 with log likelihood = -11.0712.
# Creating level 7 with log likelihood = -6.9397.
# Creating level 8 with log likelihood = -3.30518.
# Creating level 9 with log likelihood = 0.522437.
\end{verbatim}
\end{framed}
\caption{\it An example of \dnest~output that is shown on the screen. More
detailed information is saved to the three output files
{\tt sample.txt}, {\tt sample\_info.txt}, and {\tt levels.txt}.
\label{fig:output}}
\end{figure}

@@

The plots are useful for diagnosing the progress of your run. For example, you
want to ensure that the Metropolis acc
eptance fraction doesn't drop to near
zero. If it does, the proposal distributions might be too aggressive, or maybe
you just created too many levels (and the high level distributions are just
extremely narrow). The other key plot for diagnostics is the posterior
weights. On most problems, these will peak at some point, and begin to
come back down towards zero (See Figure~\ref{fig3}).
Once this occurs, you probably have enough levels.
It is wasteful to create large amounts of levels extending well beyond the peak.
However, on problems with phase transitions more than one peak may occur.
An example of not having enough levels is given in
Figure~\ref{fig:not_enough_levels}. To change the number of levels, you
can use the OPTIONS file, described in Section~\ref{sec:options}.\\

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{fig1.pdf}
\caption{\it Level vs. Time from a \dnest~run with 100 levels.\label{fig:fig1}}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{framed}
\begin{verbatim}
DNest4 Command Line Options: 
-h: display this message
-o <filename>: load DNest3 options from the specified file. Default=OPTIONS
-s <seed>: seed the random number generator with the specified value. If unspecified, the system time is used.
-d <filename>: Load data from the specified file, if required.
-c <value>: Specify a compression value (between levels) other than e.
-t <num_threads>: run on the specified number of threads. Default=1.
\end{verbatim}
\end{framed}
\end{figure}

The most important option is the one about the number of threads. Depending on
your computer, you probably want to run \dnest~with more than one thread.
Each thread will be responsible for evolving its own walker(s), and periodically
the threads will pool their results (for example, information about the
levels). To execute \dnest~with 8 threads, simply use {\tt ./main -t 8}. You can
replace 8 with however many threads you want. I usually use 8 threads because
that's how many my computer can run simultaneously at full speed.\\

@@

@@

\section{Implementing models}
\dnest~currently includes three example models: SpikeSlab, StraightLine, and
RJObject_SineWaves. It also contains an empty Template example which you can use
to implement your own models.\\

To use \dnest~on a particular problem, you need to implement a C++ class
which describes the features of the model (e.g. what the parameters are,
what the prior is, and what the likelihood is). Specifically, an object
(instance) of your class represents a point in the parameter space of your
problem. The class must define several member functions which can be
applied to any object. The first is {\tt from_prior(DNest4::RNG&)},
which when called,
initialises the object with parameter values from the prior distribution
(the argument is a random number generator object, passed by reference).
The second important member function is {\tt perturb(DNest4::RNG&)}, which is used to propose
a change to the parameters, as in the basic Metropolis algorithm. The third
is {\tt log_likelihood()}, which returns (as you might expect)
the log of the likelihood function evaluated at the object's current position. Next,
we have the {\tt print(std::ostream\&)} member function which defines how the model
parameters are to be written to the file (i.e. which parameters are to be
printed, and in what order). Finally, there is {\tt description()},
which allows you to put a header at the top of the output files with information
about what is in each column. To demonstrate how these member functions work, we'll go
through the code for {\tt SpikeSlab}.\\

In the SpikeSlab directory,
the relevant class is the {\tt SpikeSlab} class which is defined in the files
{\tt SpikeSlab.h} and {\tt SpikeSlab.cpp}. If you open {\tt SpikeSlab.h},
you'll see that the class has only one variable, which is a C++ valarray called
{\tt params}. This contains the 20 parameter values.\\

\begin{figure}
\begin{framed}
\begin{verbatim}
class SpikeSlab
{
    private:
        std::valarray<double> params;
\end{verbatim}
\end{framed}
\end{figure}

It's up to you whether you want to have a single vector or valarray of parameters, like
in {\tt SpikeSlab}, or if you want to name all your parameters with names that
make sense to you.\\

In {\tt SpikeSlab.cpp}, all of the member function of the {\tt SpikeSlab} class
are defined. First, we have the {\it constructor}, which defines what happens
when we create an object of class {\tt SpikeSlab}:\\

@@

This constructor does nothing except specify the size of the {\tt params}
valarray, which in this example is 20. Next we have {\tt from_prior(DNest4::RNG&)}, the
function which generates the 20 parameters from the prior. In the SpikeSlab
example, the prior is Uniform$(-0.5, 0.5)$, so this function just has a loop
which generates each value from this distribution. Here I'm making use of
a function {\tt rng.rand()} which generates values from $U(0, 1)$ using the
random number generator called rng (which is passed in by \dnest).
The RNG class is defined in {\tt RNG.h} within \dnest, and is really just a
wrapper of C++11's random number generator, with more convenient function names.

\begin{framed}
\begin{verbatim}
void SpikeSlab::from_prior(RNG& rng)
{
	for(size_t i=0; i<params.size(); i++)
		params[i] = -0.5 + rng.rand();
}
\end{verbatim}
\end{framed}

The next important member function is {\tt perturb(DNest4::RNG&}, which defines the Metropolis
proposals that are used to explore. Since \dnest~spends some time sampling
the prior, and some time sampling narrow distributions like the posterior,
it's a good idea to use very heavy tailed proposal distributions. In the
SpikeSlab example, the prior width is 1, so you shouldn't make moves much
bigger than that, but you don't have any real idea how narrow the posterior
distribution is until you've solved the problem. In the SpikeSlab example,
the proposal is done by choosing one of the 20 parameters to move, and making
a move according to a scale mixture of gaussians (equivalent to ``a gaussian
proposal with a random step size''). Finally, to prevent any proposals outside
the prior range, I use modulo to ensure the proposed value is between -0.5
and 0.5. The resulting proposal is:\\

\begin{framed}
\begin{verbatim}
double SpikeSlab::perturb(RNG& rng)
{
	int which = rng.rand_int(params.size());
    params[which] += pow(10., 1.5-6*rng.rand())*rng.randn();
    params[which] = mod(params[which] + 0.5, 1.) - 0.5;
    return 0.;
}
\end{verbatim}
\end{framed}

For convenience, I recently defined two useful functions {\tt randh()}
(to generate heavy tailed proposals) and
{\tt wrap(double\& value, double min, double max)} (to simplify the modulo
stuff). An equivalent version of {\tt perturb} is then:\\

\begin{framed}
\begin{verbatim}
double SpikeSlab::perturb(RNG& rng)
{
	int which = rng.rand_int(params.size());
	params[which] += rng.randh();
	wrap(params[which], -0.5, 0.5);
	return 0.;
}
\end{verbatim}
\end{framed}

The last thing we need to pay attention to in the perturb function is the
returned value. \dnest~assumes that your perturb function, if used repeatedly,
would inherently explore the prior distribution in your problem. With
a uniform prior, it's easy to ensure this using wrap, as shown above.
However, for a non-uniform prior,
it's not so easy. One way to do it is to work in a coordinate system where
the prior {\it is} uniform, and then transforming the parameter before using
it in the likelihood function. Another way is to return a nonzero value from
the {\tt perturb} function. Any value returned is added to the log of the
acceptance probability. So if you want to explore a non-uniform prior, you
can return the log of the ratio of the prior density at the proposed point
to the value at the original point. For concreteness, suppose you had a
parameter $a$ with a Uniform$(0,10)$ prior, and a parameter $b$ with a
Normal$(20, \sigma=5)$ prior. Then the {\tt perturb} member function could be implemented
like this:\newpage

\begin{framed}
\begin{verbatim}
double MyModel::perturb(RNG& rng)
{
    int which = rng.rand_int(2);
    double logH = 0.;

    if(which == 0)
    {
        a += 10*rng.randh();
        wrap(a, 0., 10.);
    }
    else
    {
        logH -= -0.5*pow((b - 20.)/5., 2);
        b += 5*rng.randh();
        logH += -0.5*pow((b - 20.)/5., 2);
    }

@@

The {\tt log_likelihood} member does what you'd expect: evaluates and returns
the log likelihood based on the current setting of the parameters. In the
SpikeSlab example, the likelihood is an invented mixture of two gaussians.
However, in general, your likelihood will depend on a data set. When I'm using
real data, I usually write a {\tt Data} class, an object of which represents
a data set. Since you need to access the dataset from within another class,
it's easiest to make the C++ equivalent of a ``global variable'': a single
static instance of the dataset which is accessible from everywhere. See
the StraightLine example if you want to see a Data class.\\

Finally, the {\tt print} function defines how your parameters are printed to the
{\tt sample.txt} (and subsequently {\tt posterior\_sample.txt}) files. In
SpikeSlab, {\tt print} writes all 20 parameters, separated by
spaces, to a single line of the stream ``out'' which is input to the function:\\

@@

Your print function needs to output parameters on a single line, in any order
you want. These will be the order in which they'll appear in the output files
{\tt sample.txt} and {\tt posterior\_sample.txt}. As an example, for our
above hypothetical model with parameters $a$ and $b$, the print function
and the description function might look like this:

@@

The Makefile in the Template directory should allow you to compile your model
by issuing the command {\tt make} in a terminal. The only preparation you need
to do is add the path to your \dnest~directory to an environment variable called
DNEST4\_PATH so that the compiler knows where \dnest~is on your system.
For example, on my laptop, \dnest~is located in /home/brewer/Projects/DNest4.
So I set DNEST4_PATH to /home/brewer/Projects.

If everything worked well then after running {\tt make},
the executable {\tt main} should exist, which you can use to run
\dnest~on your model. As always, the first time you run \dnest~on a new model,
you should pay attention to the output plots and make sure everything is okay,
such as the number of levels etc.

%\subsection{Implementing a model using Julia}
%Experimental support for models implemented in Julia is provided in the
%{\tt Examples/JuliaModel} directory. To use this, follow these steps:

%\begin{enumerate}
%\item Implement your log likelihood in {\tt julia\_model.jl}, as a function
%	called {\tt log\_likelihood} that takes a {\tt Vector::Float64} as an argument
%	and returns a {\tt Float64}. You should assume that the priors for the
%	parameters are independent Uniform(0, 1) distributions. To use
%	a different prior, transform the parameters inside
%	your log likelihood function before moving them.\\
%\item You may add other functions to {\tt julia\_model.jl}, for example to load
%	a dataset to refer to in your log likelihood.\\
%\item Edit the constructor for {\tt MyModel} in {\tt MyModel.cpp} so the parameter
%	vector has the correct length for your problem. The default given value
%	is 20.\\
%\item Open the {\tt Makefile} and make sure the {\tt -L} option passed to
%	{\tt g++} points to the
%	directory on your computer where the Julia shared library (such
%	as {\tt libjulia.so}) is stored. You might also have to configure your
%	system so that it searches for libraries here at runtime, e.g.
%	by running ldconfig as root. The \dnest~library and include files should
%	also be somewhere where the compiler will find them (or you can add them
%	to the {\tt g++} calls in the {\tt Makefile} using {\tt -I} for header
%	files and {\tt -L} for libraries).\\
%\item Open {\tt main.cpp} and edit the first argument to the function
%	{\tt jl\_init\_with\_image} to be the path to the file {\tt sys.so} associated with
%	Julia on your system.\\
%\item Compile by invoking '{\tt make}'.\\
%\item Run the executable! Note: only a single thread can be used, so you might want
%	to use the {\tt OPTIONS} file to have more than one particle. Five is a
%	reasonable default.
%\end{enumerate}
